% Acknowl.tex

I write this paper with unending thanks for my family, friends, the love, support and encouragement that they give, and the lessons they have taught me. With special thanks for my mother Donna, father Thomas, and sister Allison. With special thanks to my mother; her character and duty towards others has been critically inspirational for this work's focus on sustainability and climate change. I cannot emphasize how much she is the foundation of my views on the purpose of research and development. With special thanks to my Father, for inspiration through his work ethic, leadership, and open mind. In difficult times, he has always had faith and patience for me and I would not be in this program without him. With special thanks to my sister for her support of my education, curiosity, and maturity. I am eternally grateful for her sense of justice and duty; I would not be here without her support, encouragement, and acceptance. I write this with gratitude to my family for the life they have given me, the sacrifices they have made on my behalf, and most importantly their unconditional love. I write this with special thanks for my grandfather George for inspiring my pursuit of science, chemistry, and music. With special thanks for my grandmother Winnie for the inspiration of her love and the optimism which bring me through each challenge. With special thanks for my uncle Kevin for his inquisitive nature, friendship, optimism, and support. He has found creative ways to teach me about science and pioneering spirit even as busy as he is. With special thanks for my brother-in-law Coy for the friendship, encouragement, and inspiration he has shared with me; I greatly admire his sense of honesty, selflessness, and humor that have made him an \textit{excellent} role model for me. With special thanks for my nieces Violet and Ruby for their laughter, love, and their infectious energy, ingenuity, and optimism. Many thanks for girlfriend Madeline for her love, open ear, curiosity, support, and encouragement throughout the years. She has an unshakeably good heart and a responsibility towards others, which truly \textit{inspire} my belief in how my work can touch the lives of others. Many thanks to my best friend Andrew for his friendship, curiosity, support, and encouragement. A man with character and grit, we share a unique bond that has always made us a good team and great friends. With many thanks to the rest of my wonderful and supportive family and friends for their unconditional love and support. 

Many thanks to Karol Miaskiewicz for his consistent and wonderful friendship throughout this project. This would not be possible without the kindness, consideration, and respect that Karol has shown me and this project. Many thanks to all of the members of the Bioinformatics program. Particularly, I'd like to thank Erin Crowgey for her mentorship and support of my efforts to learn NGS bioinformatic analyses. Her drive, honesty, attitude, and beaming personality encapsulates who I've always believed researchers can be. With thanks to Rachel Marine, a great and inquisitive scientist and friend who always has insight and words of encouragement. With thank to Ryan Moore for his openness, honesty, encouraging reasoning, sense of humor, and of course for tossing ideas around together. I'd like to thank Shawn Polson for his open ear and perspective. I'd also like to thank Dr. Wu for her support and encouragement throughout the program and the many members of the Wu group for their support as well. Many thanks to Bruce, Olga, and Summer for their work in the Sequencing and Genotyping Center in support of this project.

Many thanks for my mentors, Drs. Keerthi Venkataramanan and Terry Papoutsakis, for their mentorship, support, critique, and faith throughout this project. The success that we've seen in this effort I owe to Keerthi's guidance and training. They are without a doubt the best scientific mentors I have ever had the pleasure of studying under. I look forward to every meeting that we have and it will be difficult to depart from their guidance when we conclude this project.
Many thanks for each of the the many members of the Papoutsakis lab for their friendship, support, and critique. I am very grateful for the unity of this group and for how much I enjoy coming to the lab. There is something unique about the Papoutsakis group. Each and every member is innovative, industrious, and full of excellent questions. Sharing lunch, riddles, cake, and time with this lab has been extraordinarily fulfilling. For the hard work, late nights, early autoclave cycles, shared spaces, and plenty of reasons to celebrate I am grateful for each and every member in this group. Because of this team of excellent scientists and engineers, this work was guided to success. Thank you, all.

{\centerline {\it Ad maiorem dei gloriam.}}