% This is the section of the results focused on the assembly


\chapter{Transcriptome Assembly}
The high sequencing depth achieved in this experiment suggested that even for low abundance transcripts, boundaries would be determined effectively. To this end, transcriptome assembly was the computational technique selected for its speed and dependence on both sequencing depth and complexity.\cite{108} Recall that assembly is resistant to the high variability of expression measurements in experiments with significant depth. It is also resistant to sequence-specific biases and low complexity background signals. 

However, it is known that transcriptome assembly is sensitive to other components of sequencing datasets such as residual adapters, substitution errors, and duplicate reads.\cite{108} In addition to the aforementioned quality trimming, adapters and primer-dimers were removed from sequenced reads and duplicate reads were removed from the dataset after alignment. This trimming and filtering strategy resulted in both paired and unpaired strand-specific reads aligning to the genome. As previously described, a large subset (450M, 83\%) of the reads were uniquely aligned and properly-paired according to the forward-reverse(FR) sequencing chemistry. The remaining data consisted of 74M discordantly aligned reads. Two assemblies were conducted to understand the effect of the extra unpaired or improperly-paired reads compared to an assembly of the properly-paired data alone.

A wide range of open source tools and approaches optimized for the transcriptome have been reviewed recently.\cite{108} Trinity was selected for transcriptome assembly after comparison with other existing approaches. Trinity\cite{58} is a flexible de-Bruijn graph assembler that accepts paired and unpaired reads alone or in combination. Perhaps more importantly, Trinity has strand-specific assembly options and improved support for bacterial genomes. This transcriptome assembler was optimal for this dataset and assembly optimization.


% This tex file includes information about the results of the assembly, including its optimization, and summarization using some simple statistics about the UTRs, and the number of ORFs included from the assembly. 
\input{uncurated_assembly.tex}
% This tex file includes some basic information about construction of the genome browser (should really be subsection of the 'uncurated_assembly', as a transition to the examples. 
\input{browser.tex}
% This tex file shows figures and a description that describes the background signal
\input{background_signal.tex}


%     A s s e m b l y     E x a m p l e s
\input{assembly_examples.tex}

\input{resolve.tex}


\section{First Strand-Specific Transcriptome Assembly for \itshape Clostridia}

In this chapter, a strand-specific transcriptome assembly was conducted with the high-depth sequencing dataset obtained for \textit{C. acetobutylicum} under various experimental conditions. The resulting boundaries were determined precisely due to laboratory and informatic quality controls and curation of the assembly. Significantly, these results represent the first strand-specific transcriptome assembly for the class \textit{Clostridia}. Moreover, the innovative approach described here addresses misassemblies that result from the limitations of modern sequencing technology that are frequently neglected by similar studies.

The initial assembly from the subset of properly-paired reads had maximal size, expression, and inclusion of reference protein annotations compared to the assembly from the full dataset. However, several types of misassemblies were present in the dataset that were observed statistically and with specific examples. To correct these errors, a customized genome browser was developed for integrative analysis of complexity, sequencing depth, and genome annotations. Heuristics for curation of the assembly were described and exemplified with canonical \textit{C. acetobutylicum} transcripts. In each case, the curation method produced transcript boundaries with precision comparable to previous targeted studies. The curation method was applied to the entire pSol1 megaplasmid, fixing misassembled transcripts. The curation technique deeply improved the type I error rate for transcript discovery and boundary determination. The integrative analyis was resilient to background signals of the sequencing technique, resulting in a high-quality assembly for the pSol1 megaplasmid. Re-evaluation of the final assembly showed transcriptome feature lengths in agreement with prokaryotic averages. These data clearly suggest the efficacy of this technique and the precision of the results.
